---
title: "Cancer"
output: html_document
---

##Part A: Loading Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Read file
wisc.df <- read.csv(file = "WisconsinCancer.csv", header = TRUE)
wisc.data <- as.matrix(wisc.df[,3:ncol(wisc.df)])
```

```{r}
#Check that our data matches the original when converted to 0 and 1's
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
sum(diagnosis)
```

Q1. How many observations are in this dataset?
```{r}
print(dim(wisc.df))
```

Q2. How many variables/features in the data are suffixed with _mean?
```{r}
grep(("_mean"), names(wisc.df))
length(grep(("_mean"), names(wisc.df)))
```

Q3. How many of the observations have a malignant diagnosis?

```{r}
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
sum(diagnosis)
```

## Part B: PCA

How does the mean and variance look throughout the data? Do we need to scale it?

Means:
```{r}
colMeans(wisc.data)
```

Standard Deviations:
```{r}
apply(wisc.data, 2, sd)
```

We can see that there is an extra column "X" full of "NA" that we need to remove

```{r}
  #wisc.df <- wisc.df[,1:(ncol(wisc.df)-1)] OR the way Barry showed us:
  wisc.df <- wisc.df[-ncol(wisc.df)]
  wisc.data <- wisc.data[,-ncol(wisc.data)]
```

Now that we've removed that non-sense column, let's make a PCA!!
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data)
```

One variable explains 98% of the data here, which is good but not what we want. We want a multivariate display. So we need to rescale it.

```{r}
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
summaryPCA <- summary(wisc.pr)
summaryPCA$importance[3,1]
```


Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
print(which(summaryPCA$importance[3,] >= .70))

```
PCA3 is the first with a cummulative variance above 70%

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
print(which(summaryPCA$importance[3,] >= .90))
```
We need 7 PCs before we can explain 90% of the  (scaled) data

## Interpreting PCA Results
Let's plot PC1 against PC2
```{r}
plot(wisc.pr$x[,c(1,2)], col = diagnosis+1, 
     xlab = "PC1", ylab = "PC2", pch = as.vector(wisc.df$diagnosis))
```

Now let's plot PC1 against PC3
```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[, c(1, 3)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC3")
```


```{r}
biplot(wisc.pr)
```
#Variance Explained: Scree Plot
```{r}
#Calculate variance of each component
pr.var <- wisc.pr$sdev^2
#Proportion of variance
pve <- pr.var / sum(pr.var)
#Plot variance explained for each prinicpal component
plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained", ylim = c(0,1), type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
# Plot cumulative proportion of variance explained
plot(cumsum(pve) , xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

par()
```{r}
#edit preferences to show 1 row 2 columns of plots
par(mfrow=c(1,2))

plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained", ylim = c(0,1), type = "o")

plot(cumsum(pve) , xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

## Heirarchical Clustering
Let's make a supervised(?) learning method 

```{r}
#Sclae the sic.data data: data.scaled
data.scaled <- scale(wisc.data)
#Calculate distance matrix needed for hclust
data.dist <- dist(data.scaled)
#
wisc.hclust.clusters <- hclust(data.dist)
```

```{r}
plot(wisc.hclust.clusters)
abline(h=20, col = 'red')
wisc.hclust.clusters <- cutree(wisc.hclust.clusters, h=20)

#or wisc.hsclus.clusters <- cutree(wisc.hclust, k=4)
```

Let's look at the clusters

```{r}
table(wisc.hclust.clusters)
```
this tells us there are 177 patients in cluster 1, 7 in cluster 2, 383 in cluster 3, and 2 in cluster 4. 


Now lets look at the groupings when cross tabulated with diagnosis
```{r}
table(wisc.hclust.clusters, diagnosis)
```

this shows us there in cluster 1, 12 patients had benign tumors while a majority of 165 patients had malignant cancers. Most of the malignant tumors are in cluster 1 and the majority of benign tumors are in cluster 3.



Maybe a different clustering would give us better results. Let's use kmeans to see what the computer thinks in the best clustering!

```{r}
wisc.km <- kmeans(wisc.data, centers= 2, nstart= 400)
table(wisc.km$cluster, wisc.hclust.clusters)
```

we can see that the y-axis (kmeans with 2 clusters) generally sorts the prior data so that most of the malignant patients are in kmeans$cluster1 and most of the begnign patients are in kmeans$cluster2. It's vaguely similar to our cutree, handmade clustering. 

```{r}
table(wisc.km$cluster, diagnosis)
```
wow! we have a somewhat improved method. We can mess with the nstart value to make the clustering "better"

##Section 5: Clustering on PCA results


```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
dist_hclust_pr <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(dist_hclust_pr)
plot(wisc.pr.hclust)
abline(h=20, col='red')
wisc.pr.hclust <- cutree(wisc.pr.hclust, k=9)
table(wisc.pr.hclust, diagnosis)
```

Let's plot our clusters!

```{r}
plot(wisc.pr$x[,c(1,2)], col = wisc.pr.hclust, 
     xlab = "PC1", ylab = "PC2", pch = as.vector(wisc.df$diagnosis))
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust2 <- kmeans(wisc.pr$x, centers = 2, nstart = 200)
table(wisc.pr.hclust2$cluster, diagnosis)
```

Let's plot our clusters!

```{r}
plot(wisc.pr$x[,c(1,2)], col = wisc.pr.hclust2$cluster, 
     xlab = "PC1", ylab = "PC2", pch = as.vector(wisc.df$diagnosis))
```