---
title: "Aleena_Lecture"
author: "Helena Tubb"
date: "5/24/2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("tidyverse")
```

```{r}
library(tidyverse)
```

```{r}
exp_data <- read.csv("Experimental_results.csv")
head(exp_data)
```

```{r}
exp_data <- select(exp_data, -X)
head(exp_data)
```

let's practice using the select function:

```{r}
practice <- select(exp_data, Mut_ID, Strain, Environment)
head(practice)
```

```{r}
exp_rearranged <- gather(exp_data, Generation, Frequency, H_0:H_96)
table_for_graph <- separate(exp_rearranged, Generation, into=c("H", "Time"))
table_for_graph <- select(table_for_graph, -H)
table_cleaned <- na.omit(table_for_graph)
table_cleaned$Time <- as.numeric(table_cleaned$Time)
head(table_cleaned)
```


```{r}
ggplot(table_cleaned)+geom_line(aes(x=Time, y=Frequency, group=BC_ID), alpha=0.2,colour="#000033")+facet_grid(.~Strain)+ggtitle("Barcode trajectories")+theme(plot.title = element_text(hjust = 0.5))+xlab("Time, hours")+ylab("Log10(Barcode frequency)")
```

could use package "plotly" that allows you to mouse over and annotate elements of the chart. It would show you which lines correspond to which strain. 

```{r}
mut34 <-filter(table_cleaned, table_cleaned$Mut_ID=="34")

ggplot(mut34,aes(Time, Frequency, group=BC_ID, color=BC_ID))+geom_line()+theme(legend.position="none")+facet_grid(.~Strain)+ggtitle("Mutation_34")+xlab("Time, hours") + ylab("Log10(Barcode frequency)")+theme(plot.title = element_text(hjust = 0.5))
```
Disregard the bottom lines, they are just noise.

```{r}
mut34_f <- filter(mut34, mut34$Frequency>=log(0.01))

ggplot(mut34_f,aes(Time, Frequency, group=BC_ID, color=BC_ID))+geom_line()+theme(legend.position="none")+facet_grid(.~Strain)+ggtitle("Mutation_34")+xlab("Time, hours") + ylab("Log10(Barcode frequency)")+theme(plot.title = element_text(hjust = 0.5))
```
Now lets select only strain 25361 in strain DivAnc
```{r}
BC_25361 <- filter(mut34_f, mut34_f$BC_ID=="25361", mut34_f$Strain=="DivAnc")

BC_25361_plot <- ggplot(BC_25361,aes(Time, Frequency, group=BC_ID, color=BC_ID))+geom_point()+theme(legend.position="none")+facet_grid(.~Strain)+geom_smooth(se=FALSE, method = "lm")+ggtitle("Mutation_34")+xlab("Time, hours") + ylab("Log10(Barcode frequency)")+theme(plot.title = element_text(hjust = 0.5))
```

Alternatively could save the ggplot code as an object and add "+geom_smooth..." etc on it. Saves typing.




```{r}
BC_25361_smooth <- BC_25361_plot+geom_smooth(se = FALSE, method = "lm")
BC_25361_smooth
```

```{r}
regression_model<-lm(Frequency~Time, BC_25361)
summary_data <- summary(regression_model)
summary_data
```

```{r}
# The information we are interested in is the value of Slopeand Intercept of this line:
# Let's try to access them:

# Time
Time<-summary_data$coefficients[2]
Time
```

```{r}
# Intercept:
Intercept<-summary_data$coefficients[1]
Intercept
```

```{r}
# Lets create the file:
data_header=matrix(data = NA,nrow = 1,ncol = 7)
        data_header[1]="Mut_ID"
        data_header[2]="BC_ID"
        data_header[3]="Strain"
        data_header[4]="Slope"
        data_header[5]="Intercept"
        data_header[6]="R^2"
write.table(data_header,"~/Desktop/temp/bimm_143_github/Aleena_Lecture/practice_output.csv",append = FALSE, sep = ",",eol="\n",dec=".",row.names = FALSE,col.names = FALSE)
```

```{r}
for (mut in unique(table_cleaned$Mut_ID)) {
    mut_data=filter(table_cleaned,table_cleaned$Mut_ID==paste(mut))
    #now we have all data for each mutation separately
    for (bc in unique (mut_data$BC_ID)) {
      #now we filtered data for each barcode within 1 mutation
      bc_mut_data=filter(mut_data,mut_data$BC_ID==paste(bc))
      for (strain in unique (bc_mut_data$Strain)) {
        str_bc_mut_data=filter(bc_mut_data,bc_mut_data$Strain==paste(strain))
        #only considering combinations with 3 or more data points - anything less is statistically insignificant
        if (nrow(str_bc_mut_data)>2){
          regression_model=lm(Frequency~Time,str_bc_mut_data)
          summary_data=summary(regression_model)
          #now write to the output file! Prepare the data array first
          data_output=matrix(data = NA,nrow = 1,ncol = 6)
          data_output[1]=mut
          data_output[2]=bc
          data_output[3]=strain
          #slope
          data_output[4]=summary_data$coefficients[2]
          #intercept
          data_output[5]=summary_data$coefficients[1]
          #r-squared
          data_output[6]=summary_data$r.squared
          #time to write
          write.table(data_output,"~/Desktop/temp/bimm_143_github/Aleena_Lecture/practice_output.csv",append = TRUE, sep = ",",eol="\n",dec=".",row.names = FALSE,col.names = FALSE)
      }
    }
  }
 }
```

